---
title: "Exploring taxa distribution across space and time"
description: >
  Learn how to query vaultkeepr to study the spatial and temporal
  distribution of a plant genus across a region using both modern
  vegetation plot data and fossil pollen archives.
vignette: >
  %\VignetteIndexEntry{Exploring taxa distribution across space and time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("config_quarto.R")
```

## Introduction

One of the most common tasks in palaeoecology is to reconstruct *where* and *when* a plant taxon was present. [VegVault](https://ondrejmottl.github.io/VegVault/) stores both modern vegetation plot data and fossil pollen archives in a single database, making it possible to reconstruct taxon distributions across time within a single workflow.

The core idea of vaultkeepr is to **build a plan first, then execute it**. Each function call — `get_datasets()`, `select_dataset_by_type()`, `get_samples()`, and so on — adds a step to a lazy SQL query stored in the `plan` object without touching the database. When the plan is fully specified, a single call to `extract_data()` executes the complete query and returns a tibble ready for analysis.

In this article we study the spatio-temporal distribution of the genus *Betula* (birch) across Western North America (WNA) from the present back to 8,000 calibrated years BP, combining modern `vegetation_plot` records (sourced from [BIEN](https://bien.nceas.ucsb.edu/bien/) and [sPlotOpen](https://idiv-biodiversity.de/en/splot/splotopen.html)) with `fossil_pollen_archive` records (sourced from [Neotoma-FOSSILPOL](https://hope-uib-bio.github.io/FOSSILPOL-website/)). Fossil pollen records extend the temporal window far beyond any modern survey, while modern vegetation plots provide the spatial detail and species-level precision needed to calibrate or validate palaeoecological reconstructions.

We use a small **example database** (built automatically below) that mirrors the structure and naming conventions of a real VegVault file. You can swap the path for your own VegVault download to run the same code on real data.

::: {.callout-tip}
## Working with real data
Download the full VegVault database from the [Database Access](https://ondrejmottl.github.io/VegVault/website/database_access.html) page to run this code on real global vegetation data. See the [Database Content](https://ondrejmottl.github.io/VegVault/website/database_content.html) page for an overview of coverage and statistics.
:::

```{r build-db, message = FALSE}
# Build (or reuse) the example database and return its file path
path_to_db <- source("helper_make_example_db.R")$value
```

---

## Step 1 — Open the vault

`open_vault()` creates a connection to the VegVault SQLite file and returns a `vault_pipe` object. This object is the **plan** — it carries both the live database connection and the accumulating lazy query through every subsequent step. No data is loaded into memory here; we are simply opening the vault and initialising the plan.

```{r open}
library(vaultkeepr)

plan <-
  open_vault(path = path_to_db)
```

---

## Step 2 — Add dataset metadata

`get_datasets()` extends the plan with the `Datasets` table — the central organising unit in VegVault. Each dataset has a unique ID, geographic coordinates (`coord_long`, `coord_lat`), a dataset type (e.g. `vegetation_plot`), a source type (e.g. `BIEN`, `sPlotOpen`, `Neotoma-FOSSILPOL`), and a sampling method. At this point every dataset in the database is in scope — subsequent steps narrow the selection.

```{r get-datasets}
plan <-
  plan |>
  get_datasets()
```

---

## Step 3 — Filter by dataset type

VegVault contains four dataset types:

- `vegetation_plot` — contemporary vegetation surveys (BIEN, sPlotOpen)
- `fossil_pollen_archive` — sediment-core pollen records (Neotoma-FOSSILPOL)
- `traits` — functional trait measurements (TRY)
- `gridpoints` — artificially created datasets holding abiotic (climate/soil) data, spatially linked to the other three types

For a full technical description of each dataset type, see the [Datasets](https://ondrejmottl.github.io/VegVault/website/database_structure.html#datasets) section of the VegVault Database Structure.

`select_dataset_by_type()` adds a type filter to the plan. Here we keep only the two types that carry occurrence information.

```{r filter-type}
plan <-
  plan |>
  select_dataset_by_type(
    sel_dataset_type = c(
      "vegetation_plot",
      "fossil_pollen_archive"
    )
  )
```

---

## Step 4 — Filter by geography

We extend the plan with a geographic filter: a bounding box in the Pacific Northwest (longitude −116 to −114 °W, latitude 44 to 46 °N), an area that captures both BIEN vegetation plots and Neotoma-FOSSILPOL pollen cores. The filter is pushed into the SQL engine — no data is transferred yet.

```{r filter-geo}
plan <-
  plan |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  )
```

---

## Step 5 — Add sample information

`get_samples()` extends the plan with sample-level metadata (age, sample size, sample name). In VegVault, *samples* are the primary unit of data — each sample belongs to exactly one dataset and carries a temporal label. Contemporary samples have `age = 0`; fossil pollen samples have `age` expressed in calibrated years before present (cal yr BP), where present = 1950 AD.

```{r get-samples}
plan <-
  plan |>
  get_samples()
```

---

## Step 6 — Filter by age

We add an age filter to the plan, keeping samples from the present (age = 0) back to 8,000 cal yr BP — the entire Holocene. This window brackets the period after the last major ice-sheet retreat, during which modern plant communities became established.

```{r filter-age}
plan <-
  plan |>
  select_samples_by_age(age_lim = c(0, 8000))
```

---

## Step 7 — Add taxa and harmonise to genus level

`get_taxa()` extends the plan with taxonomic occurrence data, including abundance values (`value`). Raw taxon names in VegVault come directly from primary sources and may differ between databases (e.g. *Betula pendula* in a vegetation plot vs. *Betula* in a pollen record). The `classify_to = "genus"` argument uses the pre-computed `TaxonClassification` table — built with the [{taxospace}](https://github.com/OndrejMottl/taxospace) package using the [GBIF Taxonomy Backbone](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c) — to roll all identifications up to genus level, making vegetation plot records and fossil pollen records directly comparable.

::: {.callout-warning}
`classify_to` triggers a computationally expensive join against the `TaxonClassification` table. On a large query this can be very slow. In addition, the taxonomic harmonisation is produced by an automated process: misclassifications can occur, especially for ambiguous names or taxa not covered by the GBIF backbone. Always verify results for taxa that are critical to your analysis.
:::

```{r get-taxa}
plan <-
  plan |>
  get_taxa(classify_to = "genus")
```

---

## Step 8 — Discover available taxon names

Before committing to a taxon filter, `get_taxon_names()` lets you browse the distinct taxon names present in the current query plan without executing a full `extract_data()` call. This is useful when you are unsure of the exact spelling, the harmonised genus name, or simply which taxa are represented in the filtered data.

`get_taxon_names()` does **not** modify the plan — it queries the database and returns a small tibble, leaving `plan` ready for the next step.

```{r taxon-names}
taxon_names <-
  plan |>
  get_taxon_names()

taxon_names
```

---

## Step 9 — Filter to a single genus

`select_taxa_by_name()` adds a taxon filter to the plan, retaining only rows matching the supplied name. We focus on *Betula* (birch), a genus with a well-documented Holocene range shift across the Northern Hemisphere that is captured in both modern vegetation surveys and fossil pollen records.

```{r filter-taxa}
plan <-
  plan |>
  select_taxa_by_name(sel_taxa = "Betula")
```

---

## Step 10 — Execute the plan

The plan is now fully specified. `extract_data()` executes the complete lazy query, fetches the results from the database into R memory, and returns a nested tibble — one row per dataset — with samples, community, and any additional data packed as list-columns. See [Exploring extract_data()](a04_extract_data.html) for a detailed walkthrough of both output modes. The `verbose = FALSE` argument suppresses progress messages.

```{r extract}
data_wna_genus <-
  plan |>
  extract_data(verbose = FALSE)

dplyr::glimpse(data_wna_genus)
```

---

## Full pipeline

All the steps above — building the plan and executing it — can be written as a single expression:

```{r full-pipeline, eval = FALSE}
data_wna_genus <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  select_dataset_by_type(
    sel_dataset_type = c(
      "vegetation_plot",
      "fossil_pollen_archive"
    )
  ) |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  ) |>
  get_samples() |>
  select_samples_by_age(age_lim = c(0, 8000)) |>
  get_taxa(classify_to = "genus") |>
  select_taxa_by_name(sel_taxa = "Betula") |>
  extract_data()
```

---

## Visualising the data

We bin the samples into 2,000-year intervals and count unique presences per bin and dataset type.

```{r plot, fig.width = 7, fig.height = 4, out.width = "100%", message = FALSE, fig.alt = "Map and time series of genus occurrences across western North America faceted by 2 000-year age bins and dataset type."}
#| code-fold: true
data_plot <-
  data_wna_genus |>
  dplyr::mutate(
    data_occ = purrr::map2(
      .x = data_samples,
      .y = data_community,
      .f = ~ dplyr::left_join(.x, .y, by = "sample_name")
    )
  ) |>
  dplyr::select(
    dataset_name, dataset_type, coord_long, coord_lat, data_occ
  ) |>
  tidyr::unnest(cols = data_occ) |>
  dplyr::filter(.data$Betula > 0) |>
  dplyr::distinct(
    dataset_name, dataset_type, age, coord_long, coord_lat
  ) |>
  dplyr::mutate(
    age_bin = paste0(
      floor(age / 2000) * 2,
      "\u20132",
      floor(age / 2000) * 2 + 2,
      " ka"
    )
  )

ggplot2::ggplot(
  data = data_plot,
  mapping = ggplot2::aes(x = age_bin, fill = dataset_type)
) +
  ggplot2::geom_bar(position = "dodge") +
  ggplot2::scale_fill_manual(
    values = palette_dataset_type,
    name = "Dataset type"
  ) +
  ggplot2::labs(
    title   = "Genus occurrence counts by age bin",
    x       = "Age bin",
    y       = "Number of samples with presence",
    caption = "Based on the vaultkeepr example database"
  ) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))
```

---

## Next steps

- Filter to multiple genera with `select_taxa_by_name(sel_taxa = c(...))`
- Retrieve the underlying references with `get_references()`
- Combine with climate data — see the [vegetation and climate article](a02_vegetation_climate.html)
- Explore age uncertainty for the fossil samples — see the [paleoecological uncertainty article](a06_palaeo_uncertainty.html)
- A comparable real-data example — tracking *Picea* across North America — is shown in the [VegVault Usage Examples](https://ondrejmottl.github.io/VegVault/website/database_examples.html#example-1-spatiotemporal-patterns-of-the-picea-genus-across-north-america-since-the-lgm)
- When publishing results, cite VegVault using the [data paper](https://doi.org/10.1038/s41597-025-06176-1) and use `get_references()` for primary source citations
