---
title: "Combining vegetation data with climate variables"
description: >
  Learn how to link modern vegetation plot data with abiotic climate
  variables (e.g. mean annual temperature) from co-located gridpoint
  datasets, preparing data for a species distribution model.
vignette: >
  %\VignetteIndexEntry{Combining vegetation data with climate variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("config_quarto.R")
```

## Introduction

Species distribution models (SDMs) require paired observations of species occurrences and environmental predictors at the same locations. [VegVault](https://ondrejmottl.github.io/VegVault/) stores both modern `vegetation_plot` records and climate `gridpoints` in a single database. The `get_abiotic_data()` function automatically links each vegetation sample to spatially and temporally proximate gridpoint samples, so you never have to manage the spatial join yourself.

The core idea of vaultkeepr is to **build a plan first, then execute it**. Each function call — `get_datasets()`, `select_dataset_by_type()`, `get_samples()`, and so on — adds a step to a lazy SQL query stored in the `plan` object without touching the database. When the plan is fully specified, a single call to `extract_data()` executes the complete query and returns a tibble ready for analysis.

Abiotic data in VegVault come from [CHELSA](https://chelsa-climate.org/) (high-resolution climate), [CHELSA-TRACE21](https://chelsa-climate.org/chelsa-trace21k/) (palaeoclimate back to 21 ka), and [WoSIS](https://www.isric.org/explore/wosis) (soil properties). Variable names follow the original source conventions: CHELSA bioclimatic variables are named `bio1`–`bio19`, and the WoSIS soil classification is `HWSD2`. A full description is in the [VegVault documentation](https://ondrejmottl.github.io/VegVault/website/database_structure.html#abiotic-variables).

In this article we prepare a dataset for an SDM over Europe: we extract modern (age = 0) vegetation plots together with mean annual air temperature (`bio1`, °C) from the nearest gridpoints.

We use a small **example database** (built automatically below) that mirrors the naming conventions of a real VegVault file. Swap the path for your own VegVault download to run on real data.

```{r build-db, message = FALSE}
path_to_db <- source("helper_make_example_db.R")$value
```

---

## Step 1 — Open the vault

`open_vault()` creates a connection to the VegVault SQLite file and returns a `vault_pipe` object. This object is the **plan** — it carries both the live database connection and the accumulating lazy query through every subsequent step. Nothing is loaded into memory here.

```{r open}
library(vaultkeepr)

plan <-
  open_vault(path = path_to_db)
```

---

## Step 2 — Add dataset metadata

`get_datasets()` extends the plan with the `Datasets` table — the central organising unit in VegVault. Each row is a unique dataset identified by its geographic coordinates, dataset type, source type (e.g. `BIEN`, `sPlotOpen`, `gridpoints`), and sampling method. At this point every dataset in the database is in scope.

```{r get-datasets}
plan <-
  plan |>
  get_datasets()
```

---

## Step 3 — Filter by dataset type

We need both `vegetation_plot` (the occurrence data) and `gridpoints` (the climate data). Including both types in the same query lets `get_abiotic_data()` find the correct spatial links.

```{r filter-type}
plan <-
  plan |>
  select_dataset_by_type(
    sel_dataset_type = c(
      "vegetation_plot",
      "gridpoints"
    )
  )
```

---

## Step 4 — Filter by geography

We restrict to a Central-European bounding box (longitude 14–16 °E, latitude 44–46 °N) centred on the Mediterranean–temperate transition zone. Filtering is pushed into the SQLite engine: only datasets whose coordinates fall within the bounding box are retained, so no unnecessary data is transferred.

Note that **gridpoints datasets are not filtered by geography** when `vegetation_plot` or `fossil_pollen_archive` datasets are also present: `get_abiotic_data()` handles the spatial matching internally and needs all nearby gridpoints to be available.

```{r filter-geo}
plan <-
  plan |>
  select_dataset_by_geo(
    long_lim = c(14, 16),
    lat_lim  = c(44, 46)
  )
```

---

## Step 5 — Add sample information

`get_samples()` extends the plan with sample-level metadata: age, sample name, sample size, and additional details. Samples with `age = 0` are contemporary (modern) surveys; pollen core samples carry age in calibrated years BP. Every sample belongs to exactly one dataset, but a dataset may contain multiple samples if they differ in time.

```{r get-samples}
plan <-
  plan |>
  get_samples()
```

---

## Step 6 — Restrict to modern samples only

Contemporary vegetation surveys in VegVault are assigned `age = 0` (years before present). Setting `age_lim = c(0, 0)` adds an age filter to the plan, keeping only those samples and discarding any fossil pollen or historical records that may be present in the filtered datasets.

We restrict the age filter to `"vegetation_plot"` via `sel_dataset_type` so that gridpoint samples (which cover multiple time periods) are **not** removed. `get_abiotic_data()` in the next step needs all gridpoint time slices to find the correct spatial–temporal match for each vegetation sample.

```{r filter-age}
plan <-
  plan |>
  select_samples_by_age(
    age_lim = c(0, 0),
    sel_dataset_type = "vegetation_plot"
  )
```

---

## Step 7 — Attach abiotic data

`get_abiotic_data()` extends the plan with the spatial–temporal linkage between vegetation samples and nearby gridpoint samples. Under the hood it reads the `AbioticDataReference` table, which pre-computes the distance (km and years) between every non-gridpoint sample and every gridpoint within 50 km and 5,000 years. The default `mode = "nearest"` selects the single closest gridpoint for each vegetation sample; `mode = "mean"` averages all surrounding gridpoints instead. At this stage all abiotic variables are attached — we filter to a single variable in the next step.

```{r get-abiotic, message = FALSE}
plan <-
  plan |>
  get_abiotic_data()
```

---

## Step 8 — Select mean annual air temperature (bio1)

`select_abiotic_var_by_name()` adds a variable filter to the plan, retaining only the abiotic variable(s) matching the supplied name. In real VegVault, CHELSA bioclimatic variables are named `bio1` through `bio19` following standard conventions:

| Variable | Description | Unit |
|----------|-------------|------|
| `bio1` | Mean annual air temperature | °C |
| `bio4` | Temperature seasonality (std dev × 100) | °C |
| `bio12` | Annual precipitation amount | kg m⁻² yr⁻¹ |
| `HWSD2` | Soil classification (WoSIS) | Unitless |

All eight abiotic variables available in VegVault, with full descriptions and units, are listed in the [Abiotic Variables](https://ondrejmottl.github.io/VegVault/website/database_structure.html#abiotic-variables) section of the VegVault Database Structure.

We select `"bio1"` here to pair each vegetation plot with its local mean annual air temperature.

```{r filter-abiotic}
plan <-
  plan |>
  select_abiotic_var_by_name(sel_var_name = "bio1")
```

---

## Step 9 — Add taxa

`get_taxa()` extends the plan with the `SampleTaxa` and `Taxa` tables, adding taxon names and abundance values (`value`). Here we use the default `classify_to = NULL` (no harmonisation) to work with the original species-level identification, which gives the finest-grained estimate of per-sample species richness. For cross-database comparisons you would set `classify_to = "genus"` or `classify_to = "family"`.

::: {.callout-warning}
`classify_to` triggers a computationally expensive join against the `TaxonClassification` table. On a large query this can be very slow. In addition, the taxonomic harmonisation is produced by an automated process: misclassifications can occur, especially for ambiguous names or taxa not covered by the GBIF backbone. Always verify results for taxa that are critical to your analysis.
:::

```{r get-taxa}
plan <-
  plan |>
  get_taxa()
```

---

## Step 10 — Execute the plan

The plan is now fully specified. `extract_data()` executes the complete lazy query and returns a nested tibble — one row per dataset — with sample metadata in `data_samples`, taxon abundances in `data_community` (for vegetation plots), and climate values in `data_abiotic` (linked from the matching gridpoints). See [Exploring extract_data()](a04_extract_data.html) for a detailed walkthrough of both output modes.

```{r extract}
data_eu_sdm <-
  plan |>
  extract_data(verbose = FALSE)

dplyr::glimpse(data_eu_sdm)
```

---

## Full pipeline

```{r full-pipeline, eval = FALSE}
data_eu_sdm <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  select_dataset_by_type(
    sel_dataset_type = c("vegetation_plot", "gridpoints")
  ) |>
  select_dataset_by_geo(
    long_lim = c(14, 16),
    lat_lim  = c(44, 46)
  ) |>
  get_samples() |>
  select_samples_by_age(age_lim = c(0, 0)) |>
  get_abiotic_data() |>
  select_abiotic_var_by_name(sel_var_name = "bio1") |>
  get_taxa() |>
  extract_data()
```

---

## Visualising the data

`data_eu_sdm` holds one row per dataset. Vegetation plot datasets carry taxon abundances in `data_community` (wide format: one column per taxon) and the linked climate values in `data_abiotic`. We unnest both to compute per-sample species richness and pair each sample with its mean annual air temperature.

```{r plot, fig.width = 6, fig.height = 5, out.width = "100%", message = FALSE, fig.alt = "Scatter plot of mean annual temperature versus taxa richness for vegetation plots and climate model data across Europe."}
# Species richness per sample from the nested community data
data_richness <-
  data_eu_sdm |>
  dplyr::filter(dataset_type == "vegetation_plot") |>
  dplyr::select(dataset_name, data_community) |>
  tidyr::unnest(cols = data_community) |>
  tidyr::pivot_longer(
    cols = -c(dataset_name, sample_name),
    names_to = "taxon",
    values_to = "abund"
  ) |>
  dplyr::filter(abund > 0) |>
  dplyr::group_by(dataset_name, sample_name) |>
  dplyr::summarise(
    n_taxa = dplyr::n_distinct(taxon),
    .groups = "drop"
  )

# Temperature (bio1) per sample from the nested abiotic data
data_temp <-
  data_eu_sdm |>
  dplyr::filter(dataset_type == "vegetation_plot") |>
  dplyr::select(dataset_name, data_abiotic) |>
  tidyr::unnest(cols = data_abiotic) |>
  dplyr::select(dataset_name, sample_name, abiotic_value) |>
  dplyr::distinct()

# Join richness to temperature
data_for_plot <-
  data_richness |>
  dplyr::left_join(
    data_temp,
    by = dplyr::join_by(dataset_name, sample_name)
  ) |>
  tidyr::drop_na(abiotic_value)

ggplot2::ggplot(
  data = data_for_plot,
  mapping = ggplot2::aes(x = abiotic_value, y = n_taxa)
) +
  ggplot2::geom_point(
    colour = col_blue_light,
    alpha  = 0.7,
    size   = 2
  ) +
  ggplot2::geom_smooth(
    method  = "lm",
    formula = y ~ x,
    colour  = col_brown_dark,
    se      = FALSE
  ) +
  ggplot2::labs(
    title   = "Species richness vs. mean annual air temperature (bio1)",
    x       = "Mean annual air temperature (bio1, \u00b0C)",
    y       = "Number of taxa per vegetation plot",
    caption = "Based on the vaultkeepr example database (CHELSA bio1)"
  )
```

---

## Next steps

- Select additional climate variables with `select_abiotic_var_by_id()`
- Change the spatial linking mode via `get_abiotic_data(mode = "mean")`
- Add historical climate data by relaxing the age filter
- Retrieve data citations with `get_references()`
- Explore both output modes of `extract_data()` — see the [Exploring extract_data()](a04_extract_data.html) article
- Combine with trait data — see the [functional traits article](a03_functional_traits.html)
- A comparable real-data example — an SDM for all vegetation in Czechia — is shown in the [VegVault Usage Examples](https://ondrejmottl.github.io/VegVault/website/database_examples.html#example-2-joined-species-distribution-model-for-all-vegetation-within-czechia)
