---
title: "Exploring extract_data(): from ready-to-use output to raw data"
description: >
  A focused walkthrough of the extract_data() function. Starts with the
  default nested output (return_raw_data = FALSE) — the recommended starting
  point for most users — then introduces the raw flat output
  (return_raw_data = TRUE) for advanced workflows that require speed or direct
  access to the database structure.
vignette: >
  %\VignetteIndexEntry{Exploring extract_data(): from ready-to-use output to raw data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Every vaultkeepr pipeline follows the same two-phase pattern: **build the plan, then execute it**. Each function call — `get_datasets()`, `select_dataset_by_type()`, `get_samples()`, `get_taxa()`, etc. — adds a step to a lazy SQL query stored in the `plan` object without pulling any data into R. `extract_data()` is the execution step: it runs the complete plan against the database and returns a tibble ready for analysis.

`extract_data()` offers two output modes controlled by the `return_raw_data` argument:

- **`return_raw_data = FALSE`** (default) — the result is passed through `get_readable_column_names()` (which translates internal database column names to descriptive equivalents) and `pack_data()` (which nests sample, community, trait, and abiotic data into dedicated list-columns, one row per dataset). This is the recommended output for most users.
- **`return_raw_data = TRUE`** — the raw SQL result is returned immediately, skipping all post-processing. It is faster and preserves every column exactly as it comes from the database, but requires familiarity with the [VegVault database schema](https://ondrejmottl.github.io/VegVault/website/database_structure.html).

We use a small **example database** (built automatically below) that mirrors a real VegVault file. You can swap the path for your own VegVault download to run the same code on real data.

::: {.callout-tip}
## Working with real data
Download the full VegVault database from the [Database Access](https://ondrejmottl.github.io/VegVault/website/database_access.html) page. For an overview of the extraction workflow, see [VegVault Usage Examples](https://ondrejmottl.github.io/VegVault/website/database_examples.html#data-extraction-workflow).
:::

```{r build-db, message = FALSE}
# Build (or reuse) the example database and return its file path
path_to_db <- source("helper_make_example_db.R")$value
```

---

## Building the plan

We start with a minimal plan that retrieves vegetation plot datasets from a small geographic area together with their taxa. This gives us both sample metadata and community (taxon abundance) data — enough to illustrate both output modes clearly.

```{r build-pipeline}
library(vaultkeepr)

plan <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  select_dataset_by_type(sel_dataset_type = "vegetation_plot") |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  ) |>
  get_samples() |>
  get_taxa()
```

At this point `plan` holds the complete lazy query; no data has been collected yet.

---

## Default output — Nested (packed) tibble (`return_raw_data = FALSE`)

With the default mode, `extract_data()` executes the plan and then passes the SQL result through two additional steps before returning it:

1. `get_readable_column_names()` — translates internal database column names to descriptive equivalents.
2. `pack_data()` — nests sample, community, trait, and abiotic data into dedicated list-columns, one row per dataset.

```{r extract-packed}
data_nested <-
  plan |>
  extract_data(
    return_raw_data = FALSE,
    verbose = FALSE
  )

dplyr::glimpse(data_nested)
```

### Structure of the nested tibble

Each row represents one **dataset**. The top-level metadata columns (`dataset_name`, `dataset_type`, `coord_long`, `coord_lat`, etc.) describe the dataset. Detailed data are packed into nested `data.frame` columns:

| Nested column | Content | Present when |
|---------------|---------|--------------|
| `data_samples` | Sample metadata: `sample_name`, `age`, `sample_size`, `description` | `get_samples()` was called |
| `data_community` | Community matrix: samples × taxa (wide format, taxa as columns, abundance as values) | `get_taxa()` was called |
| `data_traits` | Trait table: samples × taxa × trait values | `get_traits()` was called |
| `data_abiotic` | Abiotic table: samples linked to gridpoint variables | `get_abiotic_data()` was called |

### Working with the nested columns

Use `tidyr::unnest()` to expand a nested column back into a flat tibble, or use `purrr::map()` to operate on each dataset's data independently.

```{r unnest-samples}
# Expand sample metadata for every dataset
data_nested |>
  tidyr::unnest(cols = data_samples)
```

```{r unnest-community}
# Expand the community matrix for every dataset
data_nested |>
  tidyr::unnest(cols = data_community)
```

```{r map-richness}
# Compute per-dataset taxon richness using purrr::map()
data_nested |>
  dplyr::mutate(
    n_taxa = purrr::map_int(
      .x = data_community,
      .f = ~ {
        # Number of taxon columns = all columns except sample_name
        ncol(.x) - 1L
      }
    )
  ) |>
  dplyr::select(dataset_name, dataset_type, n_taxa)
```

### When to use the nested output

- **Recommended for most users** — the data arrives clean and clearly organised, with no schema knowledge required.
- **Iterate over datasets** with `purrr::map()`, applying the same analysis to every dataset independently.
- **Keep data organised** by dataset — useful when datasets differ in the number of samples they contain.
- **Export** each dataset's community matrix or trait table as a separate object or file.

---

## Advanced — Raw flat tibble (`return_raw_data = TRUE`)

Setting `return_raw_data = TRUE` executes the plan and returns the SQL result immediately, skipping the column-renaming and packing steps. This makes it noticeably faster on large queries.

> **Note:** This mode requires familiarity with the [VegVault database schema](https://ondrejmottl.github.io/VegVault/website/database_structure.html). Column names match the internal database fields, and mixed data types (e.g. taxon rows interspersed with abiotic rows) must be separated manually before analysis.

```{r extract-raw}
data_flat <-
  plan |>
  extract_data(
    return_raw_data = TRUE,
    verbose = FALSE
  )

dplyr::glimpse(data_flat)
```

### What the columns mean

Every row is one **sample × taxon** combination.  Key columns include:

| Column | Content |
|--------|---------|
| `dataset_id` | Unique dataset identifier |
| `dataset_type` | Type of dataset (e.g. `vegetation_plot`) |
| `coord_long` / `coord_lat` | Geographic coordinates of the dataset |
| `sample_id` | Unique sample identifier |
| `age` | Age of the sample (cal yr BP; 0 = modern) |
| `taxon_id` | Unique taxon identifier |
| `taxon_name` | Original taxon name from the source |
| `value` | Taxon abundance in the sample |

### When to use the raw output

- **Large queries** where the packing step is a bottleneck.
- Workflows that need to **join directly** on database IDs with external tables before any reshaping.
- **Debugging** or exploring the raw query result to understand what the database returned.
- Users who are **already familiar** with the VegVault schema and prefer to do their own tidying with dplyr and tidyr.

```{r raw-example}
# Count the number of samples per dataset
data_flat |>
  dplyr::distinct(dataset_id, sample_id) |>
  dplyr::count(dataset_id, name = "n_samples")
```

---

## Comparing the two outputs

Both objects represent the same plan executed in different output modes. We can verify this by checking that the number of datasets matches.

```{r compare}
# Nested: one row per dataset
nrow(data_nested)

# Flat: one row per sample × taxon combination
nrow(data_flat)

# The nested tibble covers the same datasets as the flat one
dplyr::n_distinct(data_flat$dataset_id) == nrow(data_nested)
```

---

## Full pipeline

Both modes in a single script — build the plan once, execute it twice with different output modes:

```{r full-pipeline, eval = FALSE}
# --- Default: nested (packed) output ---
data_nested <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  select_dataset_by_type(sel_dataset_type = "vegetation_plot") |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  ) |>
  get_samples() |>
  get_taxa() |>
  extract_data(return_raw_data = FALSE)

# --- Advanced: raw flat output ---
data_flat <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  select_dataset_by_type(sel_dataset_type = "vegetation_plot") |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  ) |>
  get_samples() |>
  get_taxa() |>
  extract_data(return_raw_data = TRUE)
```

---

## Next steps

- Retrieve data citations with `get_references()` before publishing results
- Combine taxa with climate data — see the [vegetation and climate article](a02_vegetation_climate.html)
- Work with trait data — see the [functional traits article](a03_functional_traits.html)
- When publishing, cite VegVault using the [data paper](https://doi.org/10.1038/s41597-025-06176-1)
