---
title: "Understanding paleoecological uncertainty"
description: >
  An introduction to the two main sources of uncertainty in paleoecological
  data stored in VegVault — age uncertainty from age-depth modelling and
  (in a future section) taxonomic uncertainty — and how to access them with
  vaultkeepr.
vignette: >
  %\VignetteIndexEntry{Understanding paleoecological uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("config_quarto.R")
```

## Introduction

Paleoecological records such as fossil pollen archives are inherently
uncertain in two distinct ways.

**Age uncertainty** arises because the age of a sediment layer is not
measured directly — it is estimated by an age-depth model (e.g.
[Bchron](https://cran.r-project.org/package=Bchron),
[OxCal](https://c14.arch.ox.ac.uk/oxcal.html)) that fits a curve
through a small number of radiocarbon dates and extrapolates between
them.  Running the model many times using a Bayesian or Monte Carlo
framework produces a distribution of plausible ages for every sample
depth — stored in VegVault as *iterations*.

**Taxonomic uncertainty** arises because pollen grains are often
identifiable only to genus or family level, because multiple species
can share the same morphotype, or because the reference collection used
for identification may differ between labs.  A future release of
vaultkeepr will expose this through a dedicated function; a placeholder
section at the end of this article describes the concept.

This article focuses on **age uncertainty**.  We build a plan for
fossil pollen archives from Western North America, then use
`get_age_uncertainty()` as a companion call to retrieve the age iterations
for every sample in scope.

We use a small **example database** (built automatically below) that
mirrors the structure and naming conventions of a real VegVault file.
Swap the path for your own VegVault download to run on real data.

::: {.callout-tip}
## Working with real data
Download the full VegVault database from the
[Database Access](https://ondrejmottl.github.io/VegVault/website/database_access.html)
page.  For an overview of available data see the
[Database Content](https://ondrejmottl.github.io/VegVault/website/database_content.html)
page.
:::

```{r build-db, message = FALSE}
# Build (or reuse) the example database and return its file path
path_to_db <- source("helper_make_example_db.R")$value
```

---

## Step 1 — Open the vault

`open_vault()` creates a connection to the VegVault SQLite file and returns a
`vault_pipe` object.  This object is the **plan** — it carries both the live
database connection and the accumulating lazy query through every subsequent
step.  No data is loaded into memory here.

```{r open}
library(vaultkeepr)

plan <-
  open_vault(path = path_to_db)
```

---

## Step 2 — Add dataset metadata

`get_datasets()` extends the plan with the `Datasets` table — the central
organising unit in VegVault.  At this point every dataset in the database is
in scope; subsequent steps narrow the selection.

```{r get-datasets}
plan <-
  plan |>
  get_datasets()
```

---

## Step 3 — Filter to fossil pollen archives

Age-depth modelling is only applied to sediment core records.  We therefore
keep only `fossil_pollen_archive` datasets — the dataset type in VegVault that
corresponds to pollen cores processed by the
[Neotoma-FOSSILPOL](https://hope-uib-bio.github.io/FOSSILPOL-website/) workflow.

```{r filter-type}
plan <-
  plan |>
  select_dataset_by_type(
    sel_dataset_type = "fossil_pollen_archive"
  )
```

---

## Step 4 — Filter by geography

We restrict to a bounding box in the Pacific Northwest (longitude −116 to
−114 °W, latitude 44 to 46 °N), an area with several well-dated pollen cores.

```{r filter-geo}
plan <-
  plan |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  )
```

---

## Step 5 — Add sample information

`get_samples()` extends the plan with sample-level metadata (`age`,
`sample_name`, `sample_size`).  For fossil pollen archives, `age` is the
**nominal** consensus age from the age-depth model (in cal yr BP, where present
= 1950 AD).  The uncertainty around this nominal age is what we retrieve in the
next section.

```{r get-samples}
plan <-
  plan |>
  get_samples()
```

---

## Step 6 — Filter by age

We keep samples from the present (age = 0) back to 8,000 cal yr BP — the
Holocene window.

```{r filter-age}
plan <-
  plan |>
  select_samples_by_age(age_lim = c(0, 8000))
```

---

## Age uncertainty

### What is stored in VegVault

For each sample in a fossil pollen archive, the age-depth model is run
multiple times.  Each run produces a slightly different age estimate for the
same sediment depth — reflecting the combined uncertainty from the radiocarbon
dates themselves, the accumulation rate model, and any tie-points used.  These
per-sample, per-iteration age estimates are stored in the `SampleUncertainty`
table with three columns:

| Column | Content |
|--------|---------|
| `sample_id` | Links back to the sample in `Samples` |
| `iteration` | Iteration number (1, 2, 3, …) |
| `age_uncertainty` | Age estimate for this iteration (cal yr BP) |

### Retrieving age uncertainties

::: {.callout-note}
`get_age_uncertainty()` is **not** a pipeline-building step.  It does not
return a `vault_pipe` and cannot be chained with `|>` like `get_datasets()` or
`get_samples()`.  Instead, it is a **companion call** on the same `plan` object
once `get_samples()` is already in the chain.  It queries the database
immediately and returns a tibble.

By default (`return_raw_data = FALSE`) the result is a **wide tibble** — one
row per sample, with `sample_name` as the identifier and one column per
age-model iteration (`iteration_1`, `iteration_2`, …).  Set
`return_raw_data = TRUE` to get the long-format tibble with columns
`sample_id`, `iteration`, and `age_uncertainty` — needed for join-based
analysis.
:::

```{r get-age-uncertainty}
data_age_unc <-
  get_age_uncertainty(con = plan)

dplyr::glimpse(data_age_unc)
```

By default the result is a **wide tibble** with one row per sample.
`sample_name` identifies the sample; columns `iteration_1` through
`iteration_5` hold the age estimate (cal yr BP) from each run of the
age-depth model.

### Visualising age uncertainty

To plot the uncertainty we need the **nominal** (consensus) age for each
sample.  We derive it as the median across all age-model iterations — a
robust consensus estimate equivalent to the published consensus age from the
age-depth model.  The code below first pivots `data_age_unc` to long format
before computing per-sample summaries.

```{r plot-age-unc, fig.width = 7, fig.height = 4, out.width = "100%", message = FALSE, fig.alt = "Scatter plot showing the 95% age-uncertainty interval width against the nominal sample age for fossil pollen archive samples from Western North America. Uncertainty grows with age."}
#| code-fold: true
# Pivot wide age uncertainty to long for per-sample analysis;
# data_age_unc_long is reused in the richness section below
data_age_unc_long <-
  data_age_unc |>
  tidyr::pivot_longer(
    cols = dplyr::starts_with("iteration_"),
    names_to = "iteration",
    values_to = "age_uncertainty"
  )

# Nominal age per sample: median across all age-model iterations
data_nominal_ages <-
  data_age_unc_long |>
  dplyr::group_by(.data$sample_name) |>
  dplyr::summarise(
    age = as.integer(round(stats::median(.data$age_uncertainty))),
    .groups = "drop"
  )

# Per-sample 95 % credible interval of the age-depth model
data_age_summary <-
  data_age_unc_long |>
  dplyr::group_by(.data$sample_name) |>
  dplyr::summarise(
    lwr_95 = stats::quantile(
      .data$age_uncertainty,
      probs = 0.025,
      na.rm = TRUE
    ),
    upr_95 = stats::quantile(
      .data$age_uncertainty,
      probs = 0.975,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  dplyr::left_join(data_nominal_ages, by = "sample_name")

ggplot2::ggplot(
  data = data_age_summary,
  mapping = ggplot2::aes(
    x    = age,
    ymin = lwr_95,
    ymax = upr_95
  )
) +
  ggplot2::geom_segment(
    mapping = ggplot2::aes(
      x = 0,
      xend = 8e3,
      y = 0,
      yend = 8e3
    ),
    colour = col_black,
    alpha = 0.5
  ) +
  ggplot2::geom_errorbar(
    colour = col_blue_light,
    width  = 100,
    alpha  = 0.7
  ) +
  ggplot2::geom_point(
    mapping = ggplot2::aes(y = age),
    colour  = col_green_dark,
    size    = 2.5
  ) +
  ggplot2::scale_x_reverse() +
  ggplot2::scale_y_reverse() +
  ggplot2::labs(
    title   = "Age-model uncertainty by nominal sample age",
    x       = "Nominal age (cal yr BP)",
    y       = "Age-model range (cal yr BP)",
    caption = paste0(
      "Error bars = 95 % credible interval across iterations. ",
      "Dots = nominal (consensus) age.\n",
      "Based on the vaultkeepr example database; all values are synthetic."
    )
  )
```

The plot shows that older samples accumulate wider uncertainty intervals — a
direct consequence of the SD term in the age-depth model scaling with age.
This pattern is typical of real Holocene pollen records, where older depths
have fewer or less precise radiocarbon constraints.

### Richness across age iterations

To show how age uncertainty propagates into an ecological metric, we compute
**genus richness per sample** under each age-model iteration.  A sample's
observed taxa never change — only the *age* assigned to it does.  Each
iteration's line places every sample at its iteration-specific age (connected
in age order), making visible how samples can shift left or right along the
time axis and altering the apparent shape of the richness curve.

::: {.callout-note}
Genus richness is fixed per sample — the observed taxa do not change across
iterations.  Only the *age* assigned to the sample changes.  `data_age_unc_long`
provides per-iteration ages keyed by `sample_name`; `data_sample_taxa` (from
the packed `extract_data()` output) provides the taxa, also keyed by
`sample_name`.  Both objects use `sample_name` as the join key, so the default
outputs from `get_age_uncertainty()` and `extract_data()` are sufficient.
:::

```{r extend-plan-taxa}
plan_taxa <-
  plan |>
  get_taxa(classify_to = "genus")
```

```{r extract-taxa, message = FALSE}
data_taxa_packed <-
  plan_taxa |>
  extract_data(verbose = FALSE)
```

```{r plot-richness-iter, fig.width = 7, fig.height = 4, out.width = "100%", message = FALSE, fig.alt = "Spaghetti plot of genus richness per sample, one thin line per age-model iteration. Each point on a line is one sample placed at its iteration-specific age. Lines widen for older samples where age uncertainty is larger."}
#| code-fold: true
# Unnest the packed community data and re-pivot to long to recover
# unique sample x genus combinations keyed by sample_name
data_sample_taxa <-
  data_taxa_packed |>
  dplyr::select("data_community") |>
  tidyr::unnest("data_community") |>
  tidyr::pivot_longer(
    cols = -"sample_name",
    names_to = "taxon_name",
    values_to = "n"
  ) |>
  dplyr::filter(!is.na(.data$n), .data$n > 0) |>
  dplyr::distinct(.data$sample_name, .data$taxon_name)

# Per-sample richness (fixed across iterations: a sample's taxa never change)
data_richness_per_sample <-
  data_sample_taxa |>
  dplyr::group_by(.data$sample_name) |>
  dplyr::summarise(
    n_taxa = dplyr::n_distinct(.data$taxon_name),
    .groups = "drop"
  )

# Attach per-iteration ages: each sample gets one row per iteration
data_richness_iter <-
  data_richness_per_sample |>
  dplyr::inner_join(data_age_unc_long, by = "sample_name") |>
  dplyr::arrange(.data$iteration, .data$age_uncertainty)

# Consensus line: each sample placed at its median age across iterations
data_richness_median <-
  data_richness_iter |>
  dplyr::group_by(.data$sample_name, .data$n_taxa) |>
  dplyr::summarise(
    age_uncertainty = stats::median(.data$age_uncertainty),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$age_uncertainty)

ggplot2::ggplot(
  data = data_richness_iter,
  mapping = ggplot2::aes(
    x = age_uncertainty,
    y = n_taxa,
    group = as.factor(.data$iteration)
  )
) +
  ggplot2::geom_line(
    colour = col_blue_light,
    alpha = 0.4,
    linewidth = 0.5
  ) +
  ggplot2::geom_line(
    data = data_richness_median,
    mapping = ggplot2::aes(group = NULL),
    colour = col_green_dark,
    linewidth = 1.2
  ) +
  ggplot2::scale_x_reverse() +
  ggplot2::labs(
    title = "Genus richness per sample across age-model iterations",
    x = "Age (cal yr BP)",
    y = "Genus richness (per sample)",
    caption = paste0(
      "Thin lines = individual age-model iterations (each sample at its ",
      "iteration-specific age, connected in age order).\n",
      "Thick line = median-age consensus. ",
      "Based on the vaultkeepr example database; all values are synthetic."
    )
  )
```

Each thin line is one age-model iteration; each point on a line is one
sample placed at the age assigned to it by that iteration.  The thick line
places each sample at its median age across all iterations as a consensus
reference.  Because age uncertainty shifts individual samples left or right,
the bundle of lines widens for older samples where radiocarbon constraints
are sparser and the age-depth model variance is larger.

---

## Full pipeline

Build the plan and retrieve the age uncertainty in a single script:

```{r full-pipeline, eval = FALSE}
plan <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  select_dataset_by_type(
    sel_dataset_type = "fossil_pollen_archive"
  ) |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  ) |>
  get_samples() |>
  select_samples_by_age(age_lim = c(0, 8000))

# Age uncertainty — default wide output (sample_name + one col per iteration)
data_age_unc <-
  get_age_uncertainty(con = plan)

# Extend with taxa for the richness-per-iteration analysis
# default packed output; unnested inside the analysis chunk below
data_taxa_packed <-
  plan |>
  get_taxa(classify_to = "genus") |>
  extract_data(verbose = FALSE)
```

---

## Next steps

- Add taxa data and explore taxon distributions — see the
  [taxa distribution article](a01_taxa_distribution.html)
- Combine with abiotic climate variables — see the
  [vegetation and climate article](a02_vegetation_climate.html)
- Retrieve all data citations before publishing — see the
  [references article](a05_references.html)
- When publishing, cite VegVault using the
  [data paper](https://doi.org/10.1038/s41597-025-06176-1)
