---
title: "Understanding paleoecological uncertainty"
description: >
  An introduction to the two main sources of uncertainty in paleoecological
  data stored in VegVault — age uncertainty from age-depth modelling and
  (in a future section) taxonomic uncertainty — and how to access them with
  vaultkeepr.
vignette: >
  %\VignetteIndexEntry{Understanding paleoecological uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("config_quarto.R")
```

## Introduction

Paleoecological records such as fossil pollen archives are inherently
uncertain in two distinct ways.

**Age uncertainty** arises because the age of a sediment layer is not
measured directly — it is estimated by an age-depth model (e.g.
[Bchron](https://cran.r-project.org/package=Bchron),
[OxCal](https://c14.arch.ox.ac.uk/oxcal.html)) that fits a curve
through a small number of radiocarbon dates and extrapolates between
them.  Running the model many times using a Bayesian or Monte Carlo
framework produces a distribution of plausible ages for every sample
depth — stored in VegVault as *iterations*.

**Taxonomic uncertainty** arises because pollen grains are often
identifiable only to genus or family level, because multiple species
can share the same morphotype, or because the reference collection used
for identification may differ between labs.  A future release of
vaultkeepr will expose this through a dedicated function; a placeholder
section at the end of this article describes the concept.

This article focuses on **age uncertainty**.  We build a plan for
fossil pollen archives from Western North America, then use
`get_age_uncertainty()` as a companion call to retrieve the age iterations
for every sample in scope.

We use a small **example database** (built automatically below) that
mirrors the structure and naming conventions of a real VegVault file.
Swap the path for your own VegVault download to run on real data.

::: {.callout-tip}
## Working with real data
Download the full VegVault database from the
[Database Access](https://ondrejmottl.github.io/VegVault/website/database_access.html)
page.  For an overview of available data see the
[Database Content](https://ondrejmottl.github.io/VegVault/website/database_content.html)
page.
:::

```{r build-db, message = FALSE}
# Build (or reuse) the example database and return its file path
path_to_db <- source("helper_make_example_db.R")$value
```

---

## Step 1 — Open the vault

`open_vault()` creates a connection to the VegVault SQLite file and returns a
`vault_pipe` object.  This object is the **plan** — it carries both the live
database connection and the accumulating lazy query through every subsequent
step.  No data is loaded into memory here.

```{r open}
library(vaultkeepr)

plan <-
  open_vault(path = path_to_db)
```

---

## Step 2 — Add dataset metadata

`get_datasets()` extends the plan with the `Datasets` table — the central
organising unit in VegVault.  At this point every dataset in the database is
in scope; subsequent steps narrow the selection.

```{r get-datasets}
plan <-
  plan |>
  get_datasets()
```

---

## Step 3 — Filter to fossil pollen archives

Age-depth modelling is only applied to sediment core records.  We therefore
keep only `fossil_pollen_archive` datasets — the dataset type in VegVault that
corresponds to pollen cores processed by the
[Neotoma-FOSSILPOL](https://hope-uib-bio.github.io/FOSSILPOL-website/) workflow.

```{r filter-type}
plan <-
  plan |>
  select_dataset_by_type(
    sel_dataset_type = "fossil_pollen_archive"
  )
```

---

## Step 4 — Filter by geography

We restrict to a bounding box in the Pacific Northwest (longitude −116 to
−114 °W, latitude 44 to 46 °N), an area with several well-dated pollen cores.

```{r filter-geo}
plan <-
  plan |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  )
```

---

## Step 5 — Add sample information

`get_samples()` extends the plan with sample-level metadata (`age`,
`sample_name`, `sample_size`).  For fossil pollen archives, `age` is the
**nominal** consensus age from the age-depth model (in cal yr BP, where present
= 1950 AD).  The uncertainty around this nominal age is what we retrieve in the
next section.

```{r get-samples}
plan <-
  plan |>
  get_samples()
```

---

## Step 6 — Filter by age

We keep samples from the present (age = 0) back to 8,000 cal yr BP — the
Holocene window.

```{r filter-age}
plan <-
  plan |>
  select_samples_by_age(age_lim = c(0, 8000))
```

---

## Age uncertainty

### What is stored in VegVault

For each sample in a fossil pollen archive, the age-depth model is run
multiple times.  Each run produces a slightly different age estimate for the
same sediment depth — reflecting the combined uncertainty from the radiocarbon
dates themselves, the accumulation rate model, and any tie-points used.  These
per-sample, per-iteration age estimates are stored in the `SampleUncertainty`
table with three columns:

| Column | Content |
|--------|---------|
| `sample_id` | Links back to the sample in `Samples` |
| `iteration` | Iteration number (1, 2, 3, …) |
| `age_uncertainty` | Age estimate for this iteration (cal yr BP) |

### Retrieving age uncertainties

::: {.callout-note}
`get_age_uncertainty()` is **not** a pipeline-building step.  It does not
return a `vault_pipe` and cannot be chained with `|>` like `get_datasets()` or
`get_samples()`.  Instead, it is a **companion call** on the same `plan` object
once `get_samples()` is already in the chain.  It queries the database
immediately and returns a plain flat tibble — one row per sample × iteration.
:::

```{r get-age-uncertainty}
data_age_unc <-
  get_age_uncertainty(con = plan)

dplyr::glimpse(data_age_unc)
```

The result contains one row for every combination of sample (in scope of
`plan`) and age-model iteration.  With five iterations per sample this tibble
is five times longer than the number of samples in the plan.

### Visualising age uncertainty

To plot the uncertainty we need the **nominal** (consensus) age for each
sample.  We derive it as the median across all age-model iterations stored
in `data_age_unc` — the median is a robust consensus estimate equivalent to
the published consensus age from the age-depth model.

```{r plot-age-unc, fig.width = 7, fig.height = 4, out.width = "100%", message = FALSE, fig.alt = "Scatter plot showing the 95% age-uncertainty interval width against the nominal sample age for fossil pollen archive samples from Western North America. Uncertainty grows with age."}
#| code-fold: true
# Nominal age per sample: median across all age-model iterations
data_nominal_ages <-
  data_age_unc |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::summarise(
    age = as.integer(round(stats::median(.data$age_uncertainty))),
    .groups = "drop"
  )

# Per-sample 95 % credible interval of the age-depth model
data_age_summary <-
  data_age_unc |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::summarise(
    lwr_95 = stats::quantile(
      .data$age_uncertainty,
      probs = 0.025,
      na.rm = TRUE
    ),
    upr_95 = stats::quantile(
      .data$age_uncertainty,
      probs = 0.975,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  dplyr::left_join(data_nominal_ages, by = "sample_id")

ggplot2::ggplot(
  data = data_age_summary,
  mapping = ggplot2::aes(
    x    = age,
    ymin = lwr_95,
    ymax = upr_95
  )
) +
  ggplot2::geom_segment(
    mapping = ggplot2::aes(
      x = 0,
      xend = 8e3,
      y = 0,
      yend = 8e3
    ),
    colour = col_black,
    alpha = 0.5
  ) +
  ggplot2::geom_errorbar(
    colour = col_blue_light,
    width  = 100,
    alpha  = 0.7
  ) +
  ggplot2::geom_point(
    mapping = ggplot2::aes(y = age),
    colour  = col_green_dark,
    size    = 2.5
  ) +
  ggplot2::scale_x_reverse() +
  ggplot2::scale_y_reverse() +
  ggplot2::labs(
    title   = "Age-model uncertainty by nominal sample age",
    x       = "Nominal age (cal yr BP)",
    y       = "Age-model range (cal yr BP)",
    caption = "Error bars = 95 % credible interval across iterations.\nDots = nominal (consensus) age."
  )
```

The plot shows that older samples accumulate wider uncertainty intervals — a
direct consequence of the SD term in the age-depth model scaling with age.
This pattern is typical of real Holocene pollen records, where older depths
have fewer or less precise radiocarbon constraints.

### Richness across age iterations

To show how age uncertainty propagates into an ecological metric, we compute
**genus richness** per 500-year age bin *under each age-model iteration*.
Each iteration assigns slightly different ages to the same samples, shifting
them between bins and changing the richness per bin accordingly.

::: {.callout-note}
This analysis requires **`return_raw_data = TRUE`** in `extract_data()`.
The default packed output translates `sample_id` to `sample_name` and drops
the raw ID, but `data_age_unc` is keyed by `sample_id`.  Retaining `sample_id`
in the raw output is the only way to join taxa presence to per-iteration age
estimates at the individual-sample level.  This is an advanced workflow;
everywhere else in these articles the default packed output is used.
:::

```{r extend-plan-taxa}
plan_taxa <-
  plan |>
  get_taxa(classify_to = "genus")
```

```{r extract-taxa, message = FALSE}
# Raw output needed to preserve sample_id for the join with data_age_unc
data_taxa_raw <-
  plan_taxa |>
  extract_data(
    return_raw_data = TRUE,
    verbose = FALSE
  )
```

```{r plot-richness-iter, fig.width = 7, fig.height = 4, out.width = "100%", message = FALSE, fig.alt = "Spaghetti plot of genus richness per 500-year age bin, one thin line per age-model iteration, with a thick mean line overlaid. Lines diverge, showing how age uncertainty shifts samples between bins."}
#| code-fold: true
# Unique sample x genus combinations
# taxon_id is already genus-level after classify_to = "genus"
data_sample_taxa <-
  data_taxa_raw |>
  dplyr::filter(!is.na(.data$taxon_id)) |>
  dplyr::distinct(.data$sample_id, .data$taxon_id)

# For each iteration, assign every sample to a 500-year age bin
# and count distinct genera in that bin
data_richness_iter <-
  data_sample_taxa |>
  dplyr::inner_join(
    data_age_unc,
    by = "sample_id",
    relationship = "many-to-many"
  ) |>
  dplyr::mutate(
    age_bin = round(.data$age_uncertainty / 500) * 500L
  ) |>
  dplyr::group_by(.data$iteration, .data$age_bin) |>
  dplyr::summarise(
    n_taxa = dplyr::n_distinct(.data$taxon_id),
    .groups = "drop"
  )

# Mean richness across iterations per bin
data_richness_mean <-
  data_richness_iter |>
  dplyr::group_by(.data$age_bin) |>
  dplyr::summarise(
    mean_n_taxa = mean(.data$n_taxa),
    .groups = "drop"
  )

ggplot2::ggplot(
  data = data_richness_iter,
  mapping = ggplot2::aes(
    x = age_bin,
    y = n_taxa,
    group = as.factor(.data$iteration)
  )
) +
  ggplot2::geom_line(
    colour = col_blue_light,
    alpha = 0.4,
    linewidth = 0.5
  ) +
  ggplot2::geom_line(
    data = data_richness_mean,
    mapping = ggplot2::aes(
      x = age_bin,
      y = mean_n_taxa,
      group = NULL
    ),
    colour = col_green_dark,
    linewidth = 1.2
  ) +
  ggplot2::scale_x_reverse() +
  ggplot2::labs(
    title = "Genus richness per age bin across age-model iterations",
    x = "Age bin (cal yr BP)",
    y = "Genus richness",
    caption = paste0(
      "Thin lines = individual age-model iterations. ",
      "Thick line = mean across iterations."
    )
  )
```

Each thin line is one age-model iteration; the thick line is the mean.
The spread between lines makes the ecological impact of age uncertainty
directly visible: samples can be re-assigned to different 500-year bins
depending on which iteration is used, producing different richness estimates
at the same nominal age.  Bins with wider uncertainty intervals (older
strata, see the plot above) show greater spread between iteration lines.

---

## Full pipeline

Build the plan and retrieve the age uncertainty in a single script:

```{r full-pipeline, eval = FALSE}
plan <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  select_dataset_by_type(
    sel_dataset_type = "fossil_pollen_archive"
  ) |>
  select_dataset_by_geo(
    long_lim = c(-116, -114),
    lat_lim  = c(44, 46)
  ) |>
  get_samples() |>
  select_samples_by_age(age_lim = c(0, 8000))

# Age uncertainty — companion call, not a pipeline step
data_age_unc <-
  get_age_uncertainty(con = plan)

# Extend with taxa for the richness-per-iteration analysis
# return_raw_data = TRUE is needed to preserve sample_id for joining
# with data_age_unc
data_taxa_raw <-
  plan |>
  get_taxa(classify_to = "genus") |>
  extract_data(
    return_raw_data = TRUE,
    verbose = FALSE
  )
```

---

## Next steps

- Add taxa data and explore taxon distributions — see the
  [taxa distribution article](a01_taxa_distribution.html)
- Combine with abiotic climate variables — see the
  [vegetation and climate article](a02_vegetation_climate.html)
- Retrieve all data citations before publishing — see the
  [references article](a05_references.html)
- When publishing, cite VegVault using the
  [data paper](https://doi.org/10.1038/s41597-025-06176-1)
