---
title: "Inspecting and customising the taxonomic classification"
description: >
  Learn how to extract the taxonomic classification table for a specific
  data call, inspect and edit it — by filtering to a single family or by
  correcting individual taxon assignments — and feed the result back into
  get_taxa() or get_traits() without re-running the full query from scratch.
vignette: >
  %\VignetteIndexEntry{Inspecting and customising the taxonomic classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("config_quarto.R")
```

## Introduction

When you call `get_taxa(classify_to = "genus")`, vaultkeepr uses the `TaxonClassification` table built automatically from the [GBIF Taxonomy Backbone](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c) via the [{taxospace}](https://github.com/OndrejMottl/taxospace) package. That automated process works well for most taxa, but you may sometimes need to:

- **inspect** which species map to which genus or family before committing to a classification;
- **restrict** the analysis to a single plant family;
- **override** the classification for a taxon that is notoriously hard to harmonise automatically.

The `get_classification_table()` function is designed for exactly this workflow. It extracts the classification table filtered to the taxa already present in your pipeline, returns it as a plain `tibble` that you can inspect and edit in R, and the result can be passed straight back to `get_taxa()` or `get_traits()` via the `classification_data` argument.

We use a small **example database** (built automatically below) that mirrors the structure and naming conventions of a real VegVault file. You can swap the path for your own VegVault download to run the same code on real data.

::: {.callout-tip}
## Working with real data
Download the full VegVault database from the [Database Access](https://ondrejmottl.github.io/VegVault/website/database_access.html) page to run this code on real global vegetation data.
:::

```{r build-db, message = FALSE}
# Build (or reuse) the example database and return its file path
path_to_db <- source("helper_make_example_db.R")$value
```

---

## Step 1 — Build the plan up to species level

We build a **base plan** that covers dataset and sample metadata but does not include taxa yet. We then branch off `plan_taxa` specifically for the classification-table inspection: `classify_to = "original"` is required because `get_classification_table()` looks up species-level `taxon_id` values in the `TaxonClassification` table. Keeping `plan_base` separate is important — it is the correct starting point for the custom and default classification pipelines in Steps 5 and 6.

```{r open}
library(vaultkeepr)

plan_base <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  get_samples()

plan_taxa <-
  plan_base |>
  get_taxa(classify_to = "original")
```

::: {.callout-note}
`plan_base` and `plan_taxa` are both `vault_pipe` objects — no data has been collected yet. The plan is assembled lazily. `get_classification_table()` in the next step is the first call that actually touches the database.
:::

---

## Step 2 — Inspect the classification table

`get_classification_table()` executes the query, filters the `TaxonClassification` table to taxa present in the plan, joins the `Taxa` name table for every level, and returns a collected `tibble` — not a `vault_pipe`. This is the "pipeline split": `plan_taxa` is kept intact for the next step.

```{r inspect-table}
data_class <-
  get_classification_table(con = plan_taxa)

dplyr::glimpse(data_class)
```

The default output (`return_raw_data = FALSE`) contains **both IDs and resolved names** for every classification level:

| Column | Content |
|---|---|
| `taxon_id` | Original species-level taxon ID |
| `taxon_name` | Species name (from `Taxa`) |
| `taxon_species` | Species taxon ID in `TaxonClassification` |
| `species_name` | Species name resolved from `taxon_species` |
| `taxon_genus` | Genus taxon ID |
| `genus_name` | Genus name |
| `taxon_family` | Family taxon ID |
| `family_name` | Family name |

The ID columns (`taxon_id`, `taxon_species`, `taxon_genus`, `taxon_family`) make the tibble directly usable as `classification_data` in subsequent calls, while the name columns let you inspect it with human-readable labels.

```{r show-table}
data_class
```

---

## Step 3 — Explore the family breakdown

A quick summary shows how many species fall into each family in the data:

```{r family-summary}
data_class |>
  dplyr::count(family_name, sort = TRUE)
```

And within a specific family:

```{r betulaceae}
data_class |>
  dplyr::filter(family_name == "Betulaceae") |>
  dplyr::select(
    taxon_name,
    genus_name,
    family_name
  )
```

---

## Step 4 — Create a restricted classification table

Suppose we want to run the genus-level analysis for **Betulaceae only** (genera *Betula*, *Alnus*, *Corylus*). We simply filter the table and pass it to `get_taxa()`.

```{r filter-class}
data_class_betulaceae <-
  data_class |>
  dplyr::filter(family_name == "Betulaceae")

data_class_betulaceae
```

::: {.callout-tip}
Any column transformation is valid here — you can also **re-map** individual species to different genera by editing the `taxon_genus` column directly, or remove rows for taxa you want to exclude entirely. The only requirement is that the table contains a `taxon_id` column and the column corresponding to `classify_to` (e.g. `taxon_genus` for `classify_to = "genus"`).
:::

---

## Step 5 — Apply the custom classification

We pass `data_class_betulaceae` back into `get_taxa()` via `classification_data`, starting from `plan_base` — the plan **before** any taxa were added. Taxa not present in the custom table are **dropped** automatically — only Betulaceae species are retained.

```{r apply-custom}
plan_betulaceae <-
  plan_base |>
  get_taxa(
    classify_to = "genus",
    classification_data = data_class_betulaceae
  )
```

---

## Step 6 — Compare default vs. custom classification

Let us execute both pipelines and compare the resulting taxa.

```{r default-result}
# Default: all families, classified to genus
data_default <-
  plan_base |>
  get_taxa(classify_to = "genus") |>
  extract_data(verbose = FALSE)
```

```{r custom-result}
# Custom: Betulaceae only
data_custom <-
  plan_betulaceae |>
  extract_data(verbose = FALSE)
```

```{r compare}
# Genera present in the default (all-families) classification
data_default |>
  dplyr::pull("data_community") |>
  purrr::map(colnames) |>
  purrr::list_c() |>
  unique() |>
  setdiff("sample_name") |>
  sort()
```

```{r compare-custom}
# Genera present in the custom (Betulaceae-only) classification
data_custom |>
  dplyr::pull("data_community") |>
  purrr::map(colnames) |>
  purrr::list_c() |>
  unique() |>
  setdiff("sample_name") |>
  sort()
```

The custom result contains only the three Betulaceae genera (*Betula*, *Alnus*, *Corylus*), while the default contains all nine genera present in the example data.

---

## Scenario B — Overriding a specific taxon assignment

Sometimes you need to **correct or re-map a single row** rather than
restrict the whole table. Typical reasons include:

- the automated GBIF harmonisation is known to misplace a taxon
  (e.g., *Senecio jacobaea* has been reclassified to *Jacobaea vulgaris*
  in many modern treatments — and automated pipelines often still
  map it back to *Senecio*);
- your analysis follows a different taxonomic authority than GBIF
  (e.g., lumping *Betula nana* with *Alnus* s.l. for broad
  palaeoecological summaries).

The edit is a two-step process:

1. Retrieve the **target genus ID** from the classification table itself.
2. Patch the affected rows with `dplyr::mutate()` and
   `dplyr::case_when()`.

```{r lookup-genus-id}
# Step 1: retrieve the taxon_genus ID for the target genus (Alnus)
id_genus_alnus <-
  data_class |>
  dplyr::filter(genus_name == "Alnus") |>
  dplyr::pull(taxon_genus) |>
  unique()

id_genus_alnus
```

```{r override-classification}
# Step 2: patch the table — reassign Betula nana to Alnus
# Only taxon_genus (the ID column) needs to change; name columns
# are display-only and have no effect on the query.
data_class_corrected <-
  data_class |>
  dplyr::mutate(
    taxon_genus = dplyr::case_when(
      taxon_name == "Betula nana" ~ id_genus_alnus,
      .default = taxon_genus
    )
  )

# Confirm: Betula nana now carries the Alnus genus ID (id_genus_alnus),
# while its genus_name display column still reads "Betula" — the
# pipeline uses only the ID, so the display value is irrelevant.
data_class_corrected |>
  dplyr::filter(
    taxon_name %in% c("Betula pendula", "Betula nana")
  ) |>
  dplyr::select(taxon_name, taxon_genus, genus_name)
```

::: {.callout-note}
Only the **ID columns** (`taxon_id`, `taxon_species`, `taxon_genus`,
`taxon_family`) influence the pipeline result. The display columns
(`genus_name`, `family_name`, …) are resolved from the database at
collection time and can be left as-is — the confirmation above shows
that *Betula nana* still reads `genus_name = "Betula"` even though
its `taxon_genus` ID now points to *Alnus*. The pipeline will group
it under *Alnus* regardless.
:::

Apply the corrected table starting from `plan_base`. Because *Betula nana*
is now lumped with *Alnus*, its abundance contributes to the `Alnus`
column instead of `Betula`.

```{r apply-corrected}
data_corrected <-
  plan_base |>
  get_taxa(
    classify_to = "genus",
    classification_data = data_class_corrected
  ) |>
  extract_data(verbose = FALSE)
```

```{r compare-corrected}
# Compare total Betula and Alnus abundance across all samples
dplyr::bind_rows(
  data_default |>
    dplyr::pull("data_community") |>
    purrr::list_rbind() |>
    dplyr::summarise(
      Betula = sum(Betula, na.rm = TRUE),
      Alnus  = sum(Alnus,  na.rm = TRUE)
    ) |>
    dplyr::mutate(classification = "default"),
  data_corrected |>
    dplyr::pull("data_community") |>
    purrr::list_rbind() |>
    dplyr::summarise(
      Betula = sum(Betula, na.rm = TRUE),
      Alnus  = sum(Alnus,  na.rm = TRUE)
    ) |>
    dplyr::mutate(classification = "corrected")
) |>
  dplyr::relocate(classification)
```

The `Betula` total is lower in the corrected result (one species removed)
and the `Alnus` total is higher by exactly the same amount (*Betula nana*
gained). Other genera are unchanged. The genus columns themselves do not
disappear — *Betula* still has four species — but the abundance budget
shifts between the two genera.

::: {.callout-tip}
To re-map **every species of a genus** rather than a single taxon, use
`genus_name == "Betula"` as the `dplyr::case_when()` condition instead
of `taxon_name == "Betula nana"`. The same approach works for family-
level remapping via `taxon_family`.
:::

---

## Using the raw-ID table

If you only need the ID columns (e.g. to edit IDs programmatically or to pass to a function that expects only the `TaxonClassification` schema), set `return_raw_data = TRUE`:

```{r raw-table}
data_class_raw <-
  get_classification_table(
    con = plan_taxa,
    return_raw_data = TRUE
  )

data_class_raw
```

This returns only `taxon_id`, `taxon_species`, `taxon_genus`, and `taxon_family`. It is slightly lighter than the default but lacks the resolved names, so it is harder to inspect by eye.

---

## Full pipeline

The complete workflow — from opening the vault to applying a custom family filter — can be written as a single expression:

```{r full-pipeline, eval = FALSE}
# Step 1: build the base plan (no taxa yet) and branch for inspection
plan_base <-
  open_vault(path = path_to_db) |>
  get_datasets() |>
  get_samples()

# Step 2: get and filter the classification table
data_class_betulaceae <-
  get_classification_table(
    con = plan_base |> get_taxa(classify_to = "original")
  ) |>
  dplyr::filter(family_name == "Betulaceae")

# Step 3: apply and extract, starting from the base plan
data_betulaceae <-
  plan_base |>
  get_taxa(
    classify_to = "genus",
    classification_data = data_class_betulaceae
  ) |>
  extract_data()
```

---

## Next steps

- Combine with climate data after restricting the taxa of interest — see the [vegetation and climate article](a02_vegetation_climate.html)
- Retrieve the full trait dataset for the restricted or overridden taxa with `get_traits(classify_to = "genus", classification_data = ...)` — the same `classification_data` argument is available there too
- Explore age uncertainty for the filtered fossil samples — see the [paleoecological uncertainty article](a06_palaeo_uncertainty.html)
- Chain both edits: first filter the table to a family, then patch individual rows within it before passing the result to `get_taxa()`
